---
title: "AGSA talk 2025"
format: html
editor: visual
---

## Library 
```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(agridat)
library(lme4)
library(mgcv)
library(emmeans)
```


## Data generated by a designed experiment 

A split-plot experiment in three blocks. Whole-plot is 'management', sub-plot is 'time' of application, with two subsamples. The data are the heights of plants, measured on two single-hill sampling units in each plot.


$$y_{ij} = \mu + G_i + r_j + \varepsilon_{ij}, \\ r_j \sim N(0\sigma^2_r), \\ \varepsilon_{ij} \sim N(0\sigma^2_\varepsilon).$$


```{r}
df_subsample <- gomez.splitplot.subsample

df_subsample |> 
  ggplot(aes(time, height))+
  geom_point(aes(color = rep))+
  facet_wrap(~manage)+
  theme_classic()
```

```{r}
m_subsample <- lmer(height ~ manage * time + (1|rep/manage/time), data = df_subsample)

VarCorr(m_subsample)
```

```{r}
DHARMa::simulateResiduals(m_subsample, plot = T)
```

## Opportunistic use of data generated by a designed experiment 

Yield data from a multienvironment trial with IBD and opportunistic weather data. 

```{r}
df_yield <- read.csv("../data/yield.csv")
df_weather <- read.csv("../data/weather.csv")

df_combined <- df_yield |> 
  left_join(df_weather)
```

```{r}
m1 <- lmer(yield ~ gen + ppt_cum_mm + temp_avg_C + county + (1|county:block),
           df_combined)
summary(m1)


m2 <- lmer(yield~ gen + ppt_cum_mm + temp_avg_C + county +
             (1|county) +
             (1|county:block),
           df_combined)
summary(m2)
```

## Mis-specified blocks 

Yield data from an advanced Nebraska Intrastate Nursery (NIN) breeding trial conducted at Alliance, Nebraska, in 1988/89.

```{r}
data <- drop_na(stroup.nin)

data |> 
  ggplot(aes(col, row))+
  geom_tile(aes(fill = rep), color = "black")+
  theme_classic()+
  labs(fill = "Block")+
  theme(axis.line = element_blank())+
  coord_equal()
```

RCBD model: 

$$y_{ij} = \mu + G_i + r_j + \varepsilon_{ij}, \\ r_j \sim N(0\sigma^2_r), \\ \varepsilon_{ij} \sim N(0\sigma^2_\varepsilon).$$

```{r}
m <- lmer(yield ~ gen + (1|rep), data = data)
data$res <- resid(m)

data %>% 
  ggplot(aes(col, row))+
  geom_tile(aes(fill = res), color = "black")+
  theme_classic()+
  scico::scale_fill_scico()+
  theme(axis.line = element_blank())+
  labs(fill = expression(Residual~(bu~ac^{-1})))+
  coord_equal()
```

```{r}
as.data.frame(emmeans(m, ~ gen)) %>% 
  arrange(-emmean) %>% 
  head()
```

```{r}
m_spatial <- gam(yield ~ gen + s(rep, bs = "re") + s(col, row, bs = "ds"), data = data)
data$resid_spatial <- resid(m_spatial)

data %>% 
  ggplot(aes(col, row))+
  geom_tile(aes(fill = resid_spatial), color = "black")+
  theme_classic()+
  coord_fixed()+
  scico::scale_fill_scico()+
  theme(axis.line = element_blank())+
  labs(fill = expression(Residual~(bu~ac^{-1})))
```

```{r}
as.data.frame(emmeans(m_spatial, ~ gen)) %>% 
  arrange(-emmean) %>% 
  head()
```


